---
title: 'Evolutionary Rescue in Conservation: A Theoretical Model inspired by the Iberian
  Lynx (Lynx pardus)'
author: "Damian Vogt"
date: "05.11.2025"
output:
  html_document:
    df_print: paged
---

### List of Parameters and Variables: 

```{r eval=FALSE, include=FALSE}
N_A <- "Mutant Pop Size "
N_a <- "Wildtype Pop Size"
N_Aa <- "Heterozygote Pop Size"
N_tot <- "Total Pop Size"
w_A <- "Intrinsic Mutant Fitness"
w_a <- "Intrinsic Wildtype Fitness"
w_Aa <- "Intrinsic Heterozygote Fitness"
sigma_env_a <- "Environmental SD for Wildtype"
sigma_env_A <- "Environmental SD for Mutant"
sigma_env_Aa <- "Environmental SD for Heterozygote"
r_a <- "Expected Wildtype Growth Rate "
r_A <- "Expected Mutant Growth Rate"
r_Aa <- "Expected Heterozygote Growth Rate"
K <- "Total Pop. Carrying Capacity"
A <- "Total Pop. Allee coefficient"
t_max <- "max. time for simulation"
max_gen <- "max. generation for simulation"
```

# 1. Defining growth functions and Simulation of first, non-panmictic generation

First step: set seed for reproducibility, then define a function to simulate the first generation.

```{r}
set.seed(42)

simulate_one_gen1 <- function(N_a, N_A, w_a, w_A,  R_rate, sigma_env_a, sigma_env_A, K, A) {
  # draw offspring according to Poisson distribution
  #N_a1 <- rpois(1, N_a * w_a *max(0,rnorm(1,mean=1,sd=sigma_env_a)))
  
  #N_A1 <- rpois(1, N_A * w_A *max(0,rnorm(1,mean=1,sd=sigma_env_A)))
  # draw new mutants according to Poisson distribution
  #mut_a_to_A <- rpois(1, offsp_a * R_rate)
  
  Ntot <- N_a + N_A # total pop
  
  dens <- max(0, 1 - Ntot / K) # density dependence on total pop. 
  
  allee <- (Ntot / (Ntot + A))^0.5  #Allee effect aff. total pop. 
  
  env_a <- max(0, rnorm(1, mean=1, sd= sigma_env_a))  #env.eff. alr normal dist.
  env_A <- max(0, rnorm(1, mean=1, sd= sigma_env_A))
  
  #r_a <- exp(w_a-(1-N_a/K)*N_a/(N_a+A))   #population growth rates
  #r_A <- exp(w_A-(1-N_A/K)*N_A/(N_A+A))
  
  r_a <- w_a*dens*allee                #corrected pop growth formula
  r_A <- w_A*dens*allee
  
  #r_a <- exp(w_a * (1 - Ntot / K)) * allee
  #r_A <- exp(w_A * (1 - Ntot / K)) * allee
  
  #N_a1 <- rpois(1, N_a*r_a*env_a)     #final step: calc. next generations
  #N_A1 <- rpois(1, N_A*r_A*env_A)
  
  N_a1 <- rpois(1, N_a*r_a*env_a)
  N_A1 <- rpois(1, N_A*r_A*env_A)
  
  # determine new population sizes of wild type and mutant
  #N_a_new <- offsp_a + N_a
  #N_A_new <-  offsp_A + N_A
  
  return(c(N_a1, N_A1))
  return(c(r_a, r_A))
}
# Test the function
print(simulate_one_gen1(100,20,1.01,1.5,0, sigma_env_a = 0.1, sigma_env_A = 0.3, K=5000, A=20))
```

# 2. Simulating non-panmictic population

Define new variables and function for population simulation

```{r}
set.seed(42)
N_init_a <- 100
N_init_A <- 20
K <- 5000
A <- 20
t_max <- 100

simulate_pop <- function(N_init_a, N_init_A, w_a, w_A, R_rate, t_max, sigma_env_a, sigma_env_A, K, A) {
  # Create the vector in which to save the results
  pop_vector <- c(N_init_a, N_init_A)
  # initiate the variables
  pop_new <- c(N_init_a, N_init_A)
  
  # run the simulation until generation t_max
  for (i in 1:t_max) {
    # redefine the current population one generation later
    pop_new <- simulate_one_gen1(pop_new[1],pop_new[2], w_a, w_A, R_rate, sigma_env_a, sigma_env_A, K, A)
    # add the new population sizes to the output vector
    pop_vector <- rbind(pop_vector,pop_new)
    # condition to stop the simulation before t_max: either the population exceeds 5000 or it goes extinct
    if (pop_new[1]+pop_new[2]==0) break
  }
  
  # define the row and column names of the output vector
  rownames(pop_vector) <- (0:t_max)[1:length(pop_vector[,1])] # note that the vector has to be cut if the simulation stopped early
  colnames(pop_vector) <- c("a","A")
  # return the result
  return(pop_vector)	
}
```

## testing the function and plotting results

```{r}

# Test the function and plot the result
# set t_max to a value to be used througout
max_gen <- 100
# create your simulation data
output <- simulate_pop(100,100,1.1,1.3,0,max_gen, sigma_env_a=0.1, sigma_env_A=0.3, 5000, 20)
# show the last few lines of the data table
print(tail(output))



# Determine x axis range
x_range <- 0:(nrow(output) - 1)

# Plot total population size
plot(x_range,
     output[,1] + output[,2],
     type = "l",
     ylim = c(0, 3000),
     xlim = c(0, max_gen),
     xlab = "Generation",
     ylab = "Population size")

# Add wild-type individuals (red)
lines(x_range, output[,1], col = "red")

# Add mutant individuals (blue)
lines(x_range, output[,2], col = "blue")

# Optional: add legend
legend("topright",
       legend = c("Total", "Wild-type (a)", "Mutant (A)"),
       col = c("black", "red", "blue"),
       lty = 1)

```

# 3.1 Extinction probabilities without rescue: varying sigma_env

```{r}
set.seed(42)
# Determine how long to run the simulation
t_max <- 20
# Determine the parameter to vary -- here it's w_a
var_sigma <- seq(0, 0.5, by = 0.025)
# compute how many values you are testing
no_var_sigma <- length(var_sigma)

# create array/data frame in which to save the results
arr <- array(NA, dim = c(no_var_sigma,2), dimnames = list(NULL,c("p_ext", "sigma_a"))) 
df <- data.frame(arr)

# # As a reminder: the function above has a bit of a too complex output
# print(simulate_pop(1000,0,1.1,1.2,0.001,tmax, sigma_env_a=0.1, sigma_env_A=0.1))
# 
# # We need only a part of the output -- a) whether or not extinction occurred and b) the frequency at the end
# # All this info is contained in the last line of the table
# tail(simulate_pop(1000,0,1.1,1.2,0.001,tmax, sigma_env_a=0.1, sigma_env_A=0.1),1)
# 
# sim_output_end <- tail(simulate_pop(1000,0,1.1,1.2,0.001,tmax, sigma_env_a=0.1, sigma_env_A=0.1),1)
# print(sim_output_end)
# 
# # this is how to compute whether extinction occurred
# out <- TRUE
# if (sum(sim_output_end)==0)
# {
# out <- FALSE
# }
# print(out)
# 
# # this is how to compute end pop size and end frequency of a
# out <- c(0,0)
# if (sum(sim_output_end)>0)
# {
# end_pop <- sum(sim_output_end)
# end_freq <- sim_output_end[1]/end_pop
# out <- c(end_pop,end_freq)
# }
# print(out)

# Parameters
N0 <- 100
K <- 5000
A <- 20

# Now let's define a function that gives us the output
run_sim_c <- function(N_init_a, N_init_A, w_a, w_A, R_rate, t_max, sigma_env_a, sigma_env_A, K, A)
{
  sim_output_end <- tail(simulate_pop(N_init_a, N_init_A, w_a, w_A, R_rate, t_max, sigma_env_a, sigma_env_A, K, A),1)
  out <- 1
  if (sum(sim_output_end)==0)
  {
    out <- 0
  }
  return(out)
}

# check <- run_sim_c(1000,0,1,1.2,0.001,tmax, sigma_env_a=0.1, sigma_env_A=0.1)
# print(check)

# Now we do this many times per parameter value 
counter <- 1
nrep <- 1000
for (choose_sig in var_sigma) {
  #set.seed(1)
  
  results <- replicate(nrep, run_sim_c(100,0,1.05,1.25,0,t_max, sigma_env_a=choose_sig, sigma_env_A=0.1, 5000, 20))
  # print(results)
  
  ###evaluation of sim_a
  p_success <- mean(results)
  p_fail <- 1-p_success
  sd <- sqrt(p_success*(1-p_success))
  
  
  df[counter,1]<- p_fail
  df[counter,2] <- choose_sig
  counter <- counter + 1
}

plot(df$sigma_a, df$p, type = "b", 
     xlab = "environmental sigma values", 
     ylab = "extinction probability")

```

# 3.2 extinction probabilities without rescue: varying w_a

```{r}
set.seed(42)
# Determine how long to run the simulation
tmax <- 20
# Determine the parameter to vary -- here it's w_a
var_w_a <- seq(0.95, 1.05, by = 0.005)
# compute how many values you are testing
no_var_w_a <- length(var_w_a)

# create array/data frame in which to save the results
arr <- array(NA, dim = c(no_var_w_a,2), dimnames = list(NULL,c("p_ext", "w_a"))) 
df <- data.frame(arr)

# # As a reminder: the function above has a bit of a too complex output
# print(simulate_pop(1000,0,1.1,1.2,0.001,tmax, sigma_env_a=0.1, sigma_env_A=0.1))
# 
# # We need only a part of the output -- a) whether or not extinction occurred and b) the frequency at the end
# # All this info is contained in the last line of the table
# tail(simulate_pop(1000,0,1.1,1.2,0.001,tmax, sigma_env_a=0.1, sigma_env_A=0.1),1)
# 
# sim_output_end <- tail(simulate_pop(1000,0,1.1,1.2,0.001,tmax, sigma_env_a=0.1, sigma_env_A=0.1),1)
# print(sim_output_end)
# 
# # this is how to compute whether extinction occurred
# out <- TRUE
# if (sum(sim_output_end)==0)
# {
# out <- FALSE
# }
# print(out)
# 
# # this is how to compute end pop size and end frequency of a
# out <- c(0,0)
# if (sum(sim_output_end)>0)
# {
# end_pop <- sum(sim_output_end)
# end_freq <- sim_output_end[1]/end_pop
# out <- c(end_pop,end_freq)
# }
# print(out)

# Parameters
N0 <- 100
K <- 5000
A <- 20

# Now let's define a function that gives us the output
run_sim_c <- function(N_init_a, N_init_A, w_a, w_A, R_rate, t_max, sigma_env_a, sigma_env_A, K, A)
{
  sim_output_end <- tail(simulate_pop(N_init_a, N_init_A, w_a, w_A, R_rate, t_max, sigma_env_a, sigma_env_A, K, A),1)
  out <- 1
  if (sum(sim_output_end)==0)
  {
    out <- 0
  }
  return(out)
}

# check <- run_sim_c(1000,0,1,1.2,0.001,tmax, sigma_env_a=0.1, sigma_env_A=0.1)
# print(check)

# Now we do this many times per parameter value 
counter <- 1
nrep <- 1000
for (choose_w_a in var_w_a) {
  #set.seed(1)
  
  results <- replicate(nrep, run_sim_c(100,0,w_a=choose_w_a,1.25,0,tmax, sigma_env_a= 0.3, sigma_env_A=0, K, A))
  # print(results)
  
  ###evaluation of sim
  p_success <- mean(results)
  p_fail <- 1-p_success
  sd <- sqrt(p_success*(1-p_success))
  
  
  df[counter,1]<- p_fail
  df[counter,2] <- choose_w_a
  counter <- counter + 1
}

plot(df$w_a, df$p_ext, type = "b", 
     xlab = "intrinsic wildtype fitness (w_a)", 
     ylab =  "extinction probability")
 

```

# 4. Simulating the population as panmictic with diploid loci

Now we want to layer our model and make the population diploid and panmictic to make it biologically more realistic. For this, we define new, updated functions and introduce new parameters and variables for the new heterozygous Genotype.

```{r}
### how to make the population diploid? 
set.seed(18)

simulation_diploid <- function(N_aa, N_Aa, N_AA, w_a, w_Aa, w_A, R_rate, t_max,
                               sigma_env_a, sigma_env_A, sigma_env_Aa, K, A)
   
  {
  
### define parameters: 
AA <- N_AA
Aa <- N_Aa
aa <- N_aa
w_aa <- 1.1
w_Aa <- 1.25
w_AA <- 1.5
sig_a <- 0.1
sig_Aa <- 0.2
sig_A <- 0.3
Ntot <- N_aa+N_Aa+N_AA
K <- 5000
A <- 20

## define dens value and allee effect
dens <- max(0, 1 - Ntot / K) # densitydependence on total pop. 
allee <- (Ntot / (Ntot + A))^0.5  #Allee effect aff. total pop. 

## define r values
r_a <- w_a*dens*allee               #corrected pop growth formula
r_A <- w_A*dens*allee
r_Aa <- w_Aa*dens*allee

## draw env values
env_a <- max(0, rnorm(1, mean=1, sd= sig_a))  #env.eff. alr normal dist.
env_A <- max(0, rnorm(1, mean=1, sd= sig_A))
env_Aa <- max(0, rnorm(1, mean=1, sd= sig_Aa))

## draw offspring for A and a from Poisson distribution
N_aa <- rpois(1, N_aa*r_a*env_a)
N_AA <- rpois(1, N_AA*r_A*env_A)
N_Aa <- rpois(1, N_Aa*r_Aa*env_Aa)

## calculate total offspring
total_off <- N_aa+N_Aa+N_AA

if (total_off==0) {
  return(c(N_aa=0, N_Aa=0, N_AA=0))
}
  #break

## calculate A and a Alleles
count_A <- N_AA*2+N_Aa
count_a <- N_aa*2+N_Aa

# calculate HW Allele frequencies
pA <- count_A/2*total_off
pa <- count_a/2*total_off

# calculate HW genotype frequencies
probs <- c(pAA= pA^2, pAa= 2*pA*pa, paa= pa^2)



## draw random offspring 
next_gen_diploid <- as.vector(rmultinom(1, size= total_off, prob=probs))
#as.numeric: returns 3x1 matrix drawn from rmultinom as a flat numeric vector
#rmultinom: assigns random value to each of three genotypes according to HW allele probabilities
#like binomial, only for multiple variables

# calculate next gen offspring
N_AA <- next_gen_diploid[1]
N_Aa <- next_gen_diploid[2]
N_aa <- next_gen_diploid[3]

c(pAA=N_AA, pAa=N_Aa, paa=N_aa)
}

simulation_diploid(100, 0, 20, 1.1, 1.2, 1.3, 0, 100, 0.1 ,0.3, 0.2, 5000, 20)

## now we simulate the diploid population
simulate_diploid <- function(N_init_a, N_init_A, N_init_Aa, w_a, w_A, w_Aa, max_gen, t_max, sig_a, sig_A, sig_Aa, K, A) {
  # Create the vector in which to save the results
  pop_vector <- c(N_init_a, N_init_A, N_init_Aa)
  # initiate the variables
  pop_new <- c(N_init_a, N_init_A, N_init_Aa)
  
  # run the simulation until generation t_max
  for (i in 1:t_max) {
    # redefine the current population one generation later
    pop_new <- simulation_diploid(pop_new[1],pop_new[2],pop_new[3], w_a, w_A, w_Aa, R_rate, sig_a, sig_A, sig_Aa, K, A)
    # add the new population sizes to the output vector
    pop_vector <- rbind(pop_vector,pop_new)
    # condition to stop the simulation before t_max: either the population exceeds 5000 or it goes extinct
    if (pop_new[1]+pop_new[2]+pop_new[3]==0) break
  }
  
  # define the row and column names of the output vector
  rownames(pop_vector) <- (0:t_max)[1:length(pop_vector[,1])] # note that the vector has to be cut if the simulation stopped early
  colnames(pop_vector) <- c("AA","Aa", "aa")
  # return the result
  return(pop_vector)	
}


# Test the function and plot the result
# set t_max to a value to be used througout
max_gen <- 100
# create your simulation data
output<- simulate_diploid(100, 20, 0, 1, 1.5, 1.25, 0, max_gen, 0.1, 0.3, sig_Aa=0.2, K=5000, A=20)
# show the last few lines of the data table
print(tail(output))



# Determine x axis range
x_range <- 0:(nrow(output) - 1)

# Plot total population size
plot(x_range,
     output[,1] + output[,2] + output[,3],
     type = "l",
     ylim = c(0, 2500),
     xlim = c(0, max_gen),
     xlab = "Generation",
     ylab = "Population size")

# Add wild-type individuals (red)
lines(x_range, output[,1], col = "red")

# Add mutant individuals (blue)
lines(x_range, output[,2], col = "blue")

# Add heterozygote (green)
lines(x_range,output[,3], col = "green")

# Optional: add legend
legend("topright",
       legend = c("Total", "Wild-type (aa)", "Mutant (AA)", "Heterozygote (Aa)"),
       title = "diploid Lynx pardus population rescue",
       col = c("black", "red", "blue", "green"),
       lty = 1)

```

## 4.1 What happens if the mutation is heterozygous?

Now we want to extend our newly diploid model even further: We assume the possibility of Captive-bred mutant individuals as heterozygotes instead of homozygotes: This means, Hetero- and Homozygous Mutants have the same Fitness and environmental sensitivities, and the Mutant allele is now dominant over the wildtype ones. How do the three resulting genotypes get fixed in this scenario, and what are the differences to the previous model?

For this scenario, we assign the fitness and environmental sigma values of the Homozygous mutants from above, to the Heterozygotes. There is no more intermediate Fitness or sigma_env values for Heterozygotes.

Other than that, the code stays identical to the previous. Now we just twitch a few parameters and reassign starting conditions and parameters to fit our assumption.

```{r}
set.seed(1)
simulation_heterozyg <- function(N_aa, N_Aa, N_AA, w_a, w_Aa, R_rate, t_max,
                               sigma_env_a, sigma_env_Aa, K, A)
  
{
  
  ### define parameters: 
  AA <- N_AA
  Aa <- N_Aa
  aa <- N_aa
  w_aa <- 1
  w_Aa <- 1.5
  
  sig_a <- 0.1
  sig_Aa <- 0.3
  
  
  Ntot <- N_aa+N_Aa+N_AA
  K <- 5000
  A <- 20
  
  ## define dens value and allee effect
  dens <- max(0, 1 - Ntot / K) # densitydependence on total pop. 
  allee <- (Ntot / (Ntot + A))^0.5  #Allee effect aff. total pop. 
  
  ## define r values
  r_a <- w_a*dens*allee               #corrected pop growth formula

  r_Aa <- w_Aa*dens*allee
  
  ## draw env values
  env_a <- max(0, rnorm(1, mean=1, sd= sig_a))  #env.eff. alr normal dist.
  
  env_Aa <- max(0, rnorm(1, mean=1, sd= sig_Aa))
  
  ## draw offspring for A and a from Poisson distribution
  N_aa <- rpois(1, N_aa*r_a*env_a)
  N_AA <- rpois(1, N_AA*r_Aa*env_Aa)
  N_Aa <- rpois(1, N_Aa*r_Aa*env_Aa)
  
  ## calculate total offspring
  total_off <- N_aa+N_Aa+N_AA
  
  if (total_off==0) {
    return(c(N_aa=0, N_Aa=0, N_AA=0))
  }
  #break
  
  ## calculate A and a Alleles
  count_A <- N_AA*2+N_Aa
  count_a <- N_aa*2+N_Aa
  
  # calculate HW Allele frequencies
  pA <- count_A/2*total_off
  pa <- count_a/2*total_off
  
  # calculate HW genotype frequencies
  probs <- c(pAA= pA^2, pAa= 2*pA*pa, paa= pa^2)
  
  
  
  ## draw random offspring 
  next_gen_diploid <- as.vector(rmultinom(1, size= total_off, prob=probs))
  #as.numeric: returns 3x1 matrix drawn from rmultinom as a flat numeric vector
  #rmultinom: assigns random value to each of three genotypes according to HW allele probabilities
  #like binomial, only for multiple variables
  
  # calculate next gen offspring
  N_AA <- next_gen_diploid[1]
  N_Aa <- next_gen_diploid[2]
  N_aa <- next_gen_diploid[3]
  
  c(pAA=N_AA, pAa=N_Aa, paa=N_aa)
}

simulation_heterozyg(100, 20, 0, 1, 1.5, 0, 100, 0.1 ,0.3, 5000, 20)

## now we simulate the diploid population
simulate_hz <- function(N_init_a, N_init_A, N_init_Aa, w_a, w_Aa, max_gen, t_max, sig_a, sig_Aa, K, A) {
  # Create the vector in which to save the results
  pop_vector <- c(N_init_a, N_init_A, N_init_Aa)
  # initiate the variables
  pop_new <- c(N_init_a, N_init_A, N_init_Aa)
  
  # run the simulation until generation t_max
  for (i in 1:t_max) {
    # redefine the current population one generation later
    pop_new <- simulation_heterozyg(pop_new[1],pop_new[2],pop_new[3], w_a, w_Aa, R_rate, sig_a, sig_Aa, K, A)
    # add the new population sizes to the output vector
    pop_vector <- rbind(pop_vector,pop_new)
    # condition to stop the simulation before t_max: either the population exceeds 5000 or it goes extinct
    if (pop_new[1]+pop_new[2]+pop_new[3]==0) break
  }
  
  # define the row and column names of the output vector
  rownames(pop_vector) <- (0:t_max)[1:length(pop_vector[,1])] # note that the vector has to be cut if the simulation stopped early
  colnames(pop_vector) <- c("AA","Aa", "aa")
  # return the result
  return(pop_vector)	
}


# Test the function and plot the result
# set t_max to a value to be used througout
max_gen <- 100
# create your simulation data
output<- simulate_hz(100, 0, 20, 1, 1.5, 0, max_gen, 0.1, sig_Aa=0.3, K=5000, A=20)
# show the last few lines of the data table
print(tail(output))



# Determine x axis range
x_range <- 0:(nrow(output) - 1)

# Plot total population size
plot(x_range,
     output[,1] + output[,2] + output[,3],
     type = "l",
     ylim = c(0, 5000),
     xlim = c(0, max_gen),
     xlab = "Generation",
     ylab = "Population size")

# Add wild-type individuals (red)
lines(x_range, output[,1], col = "red")

# Add mutant individuals (blue)
lines(x_range, output[,3], col = "blue")

# Add heterozygote (green)
lines(x_range,output[,2], col = "green")

# Optional: add legend
legend("topright",
       legend = c("Total", "Wild-type (aa)", "Mutant (AA)", "Heterozygote (Aa)"),
       title = "diploid Lynx pardus population rescue",
       col = c("black", "red", "blue", "green"),
       lty = 1)

```

# 5. What is the minimal release fraction of mutant individuals to reach target rescue probability of p\>=0.8?

Finally, we want to model for which minimal release number the probability of rescue reaches at least 80%. This is not a predictive result for real application, just an illustrative, quantitative result for this whole theoretical framework.

In this scenario, we assume fitnesses as in the first case simulation of diploid population. AA and Aa have distinct parameters in fitness and environment sensitivity, but to fit the latter model, mutant introductions always have a 50/50 ratio of AA and Aa releases, with Aa getting an edge when numbers are odd (so for instance; when 5 individuals are introduced to the population, we get 3 Aa and 2 AA's)

The initial wildtype population of (aa) to rescue is always N=100, so the more mutants are added, the bigger the initial starting population is, of course increasing survival probability. We want to find the "sweetspot" of minimum 80% rescue chance and the least necessary amount of mutants needed for release.

```{r}

set.seed(1)

### --- One-generation simulation --- ###
simulation_diploid <- function(N_aa, N_Aa, N_AA,
                               w_aa, w_Aa, w_AA,
                               sigma_env_aa, sigma_env_Aa, sigma_env_AA,
                               K, A) {
  
  Ntot <- N_aa + N_Aa + N_AA
  if (Ntot <= 0) return(c(AA = 0, Aa = 0, aa = 0))
  
  # Density dependence & Allee effect
  dens  <- max(0, 1 - Ntot / K)
  allee <- (Ntot / (Ntot + A))^0.5
  
  # Environmental noise
  env_aa <- max(0, rnorm(1, mean = 1, sd = sigma_env_aa))
  env_Aa <- max(0, rnorm(1, mean = 1, sd = sigma_env_Aa))
  env_AA <- max(0, rnorm(1, mean = 1, sd = sigma_env_AA))
  
  # Growth rates
  r_aa <- w_aa * dens * allee
  r_Aa <- w_Aa * dens * allee
  r_AA <- w_AA * dens * allee
  
  # Offspring numbers
  N_aa <- rpois(1, N_aa * r_aa * env_aa)
  N_Aa <- rpois(1, N_Aa * r_Aa * env_Aa)
  N_AA <- rpois(1, N_AA * r_AA * env_AA)
  
  total_off <- N_aa + N_Aa + N_AA
  if (total_off == 0) return(c(AA = 0, Aa = 0, aa = 0))
  
  # Allele frequencies
  pA <- (2 * N_AA + N_Aa) / (2 * total_off)
  pa <- 1 - pA
  
  # Genotype frequencies (Hardy-Weinberg)
  probs <- c(AA = pA^2, Aa = 2 * pA * pa, aa = pa^2)
  
  # Next generation (multinomial draw)
  next_gen <- as.vector(rmultinom(1, size = total_off, prob = probs))
  names(next_gen) <- c("AA", "Aa", "aa")
  
  return(next_gen)
}


### --- Full population trajectory --- ###
simulate_population <- function(N_aa0, N_Aa0, N_AA0,
                                w_aa, w_Aa, w_AA,
                                sigma_env_aa, sigma_env_Aa, sigma_env_AA,
                                K, A, t_max = 100) {
  
  pop <- matrix(NA, nrow = t_max, ncol = 3)
  colnames(pop) <- c("AA", "Aa", "aa")
  pop[1, ] <- c(N_AA0, N_Aa0, N_aa0)
  
  for (t in 2:t_max) {
    pop[t, ] <- simulation_diploid(pop[t-1, "aa"], pop[t-1, "Aa"], pop[t-1, "AA"],
                                   w_aa, w_Aa, w_AA,
                                   sigma_env_aa, sigma_env_Aa, sigma_env_AA,
                                   K, A)
    if (sum(pop[t, ]) == 0) break  # extinction
  }
  return(pop[1:t, ])
}


### --- Monte Carlo estimator --- ###
rescue_prob <- function(mutants, N_aa0 = 100,
                        w_aa = 1.1, w_Aa = 1.2, w_AA = 1.3,
                        sigma_env_aa = 0.1, sigma_env_Aa = 0.2, sigma_env_AA = 0.3,
                        K = 5000, A = 20, t_max = 100,
                        nsim = 500, rescue_threshold = 500) {
  
  N_AA0 <- floor(mutants / 2)
  N_Aa0 <- ceiling(mutants / 2)
  
  hits <- replicate(nsim, {
    out <- simulate_population(N_aa0, N_Aa0, N_AA0,
                               w_aa, w_Aa, w_AA,
                               sigma_env_aa, sigma_env_Aa, sigma_env_AA,
                               K, A, t_max)
    any(rowSums(out) >= rescue_threshold)
  })
  mean(hits)
}


### --- Sweep over introductions --- ###
target <- 0.8
mutant_range <- seq(0, 100, by = 5)

res_tab <- data.frame(mutants = mutant_range, p_rescue = NA_real_)

for (i in seq_along(mutant_range)) {
  res_tab$p_rescue[i] <- rescue_prob(mutants = mutant_range[i])
  cat("Mutants:", mutant_range[i], "- P_rescue:", round(res_tab$p_rescue[i], 3), "\n")
}

# Find minimal mutants achieving target probability
cand <- subset(res_tab, p_rescue >= target)
if (nrow(cand) > 0) {
  min_mutants <- min(cand$mutants)
  message(sprintf("Minimal number of mutants for ≥ %.0f%% rescue: %d individuals",
                  target * 100, min_mutants))
} else {
  message("No value reached the 80% rescue threshold. Increase mutant range or adjust parameters.")
}

# Plot results
plot(res_tab$mutants, res_tab$p_rescue, type = "b", pch = 16,
     xlab = "Number of introduced mutants (Aa + AA)",
     ylab = "Rescue probability",
     main = "Rescue probability vs mutant introductions")
abline(h = target, col = "red", lty = 2)

```
